---
title: 'Interesting R packages'
date: '2023-03-17'
categories:
  - R
---

## Short notes about R packages
* Flow charts in R: [ggflowchart](https://nrennie.rbind.io/blog/introducing-ggflowchart/)
* Web scraping: [rvest](https://nrennie.rbind.io/blog/web-scraping-rvest-london-marathon/)
* Spatial data / maps: [Recommendations from Nicola Rennie's blog](https://nrennie.rbind.io/blog/2022-12-17-r-packages-for-visualising-spatial-data/)

## Check package names
```{r}
#| eval: false
library('available')
# help(package = 'available')

options(available.browse = FALSE)
available('merge', )
available('track')
available('trackChanges')
available('trackR')
available('collab')
available('writeTogether')
```


## R packages built from a single *.RMD file
[Link](https://github.com/ThinkR-open/fusen)

## Interfaces to computer algebra systems in R
Herre, I have collected a few resources in the field of computer algebra systems that have an interface to `R`.


### General
* CRAN task view: Section "Multi-Precision Arithmetic and Symbolic Mathematics" on [this](https://cran.r-hub.io/web/views/NumericalMathematics.html) site
* [Blog article](https://blog.ephorie.de/doing-maths-symbolically-r-as-a-computer-algebra-system-cas)

### Maxima
* [Rim package](https://rcst.github.io/rim/)
  * extended search path according to [these](https://techpp.com/2021/09/08/set-path-variable-in-macos-guide/) instructions
  * posted issue on rim package github site because it is still not working [here](https://github.com/rcst/rim/issues/30)
  * edited /etc/launch.d
  * added shell script /usr/bin/maxima

```{r}
#| eval: false
system('export PATH=${PATH}:/Applications/Maxima.app/Contents/Resources/opt/bin')
system('echo $PATH')
help(package='rim')
```

```{r}
library(rim)
rim::maxima.isInstalled()
rim::maxima.get("1+1;")
```

```{maxima}
lambda_S: beta * psi;
Lambda_S: integrate(lambda_S, t);
S_S: exp(-Lambda_S);
assume(beta > 0);
assume(psi > 0);
E_S: integrate(S_S, t, 0, inf);
tex(''E_S);
```


### Ryacas
```{r}
library(Ryacas)
as_r(yac_str("Simplify(a*b*a^2/b-a^3)"))
yac_str('Simplify(x^2 + 2*(x-3)^3)')
yac_str("TeXForm(x^2 - 1)")
```


## Flowcharts
Source: https://cran.r-project.org/web/packages/Gmisc/vignettes/Grid-based_flowcharts.html

```{r}
library(Gmisc, quietly = TRUE)
library(glue)
library(htmlTable)
library(grid)
library(magrittr)

org_cohort <- boxGrob(glue("Stockholm population",
                           "n = {pop}",
                           pop = txtInt(1632798),
                           .sep = "\n"))
eligible <- boxGrob(glue("Eligible",
                          "n = {pop}",
                           pop = txtInt(10032),
                           .sep = "\n"))
included <- boxGrob(glue("Randomized",
                         "n = {incl}",
                         incl = txtInt(122),
                         .sep = "\n"))
grp_a <- boxGrob(glue("Treatment A",
                      "n = {recr}",
                      recr = txtInt(43),
                      .sep = "\n"))

grp_b <- boxGrob(glue("Treatment B",
                      "n = {recr}",
                      recr = txtInt(122 - 43 - 30),
                      .sep = "\n"))

excluded <- boxGrob(glue("Excluded (n = {tot}):",
                         " - not interested: {uninterested}",
                         " - contra-indicated: {contra}",
                         tot = 30,
                         uninterested = 12,
                         contra = 30 - 12,
                         .sep = "\n"),
                    just = "left")

grid.newpage()
vert <- spreadVertical(org_cohort,
                       eligible = eligible,
                       included = included,
                       grps = grp_a)
grps <- alignVertical(reference = vert$grps,
                      grp_a, grp_b) %>%
  spreadHorizontal()
vert$grps <- NULL

excluded <- moveBox(excluded,
                    x = .8,
                    y = coords(vert$included)$top + distance(vert$eligible, vert$included, half = TRUE, center = FALSE))

for (i in 1:(length(vert) - 1)) {
  connectGrob(vert[[i]], vert[[i + 1]], type = "vert") %>%
    print
}
connectGrob(vert$included, grps[[1]], type = "N")
connectGrob(vert$included, grps[[2]], type = "N")

connectGrob(vert$eligible, excluded, type = "L")

# Print boxes
vert
grps
excluded
```

## Leaflet for maps
* [Link](https://cran.r-project.org/package=leaflet)

## Here for finding the path to the project directory
* [Link](https://github.com/r-lib/here)

## Package for formatting labels in ggplot
* [Link](https://cran.r-project.org/package=ggtext)
* [Video demonstrating the package](https://posit.co/resources/videos/spruce-up-your-ggplot2-visualizations-with-formatted-text/)

## Concattenating text
* [Package glue](https://cran.r-project.org/package=glue)

## Handle missing data
* [Package nania - getting started](https://naniar.njtierney.com/articles/getting-started-w-naniar.html)
  * However, there seem to be no methods for categorical data... &rarr; [Opened issue](https://github.com/njtierney/naniar/issues/336)

```{r}
library('naniar')
library('tidyverse')
library('simputation')
aq <- airquality |> 
  mutate(Ozone = as.double(Ozone), 
         fct = sample(c('a', 'b', NA), replace = TRUE, size = nrow(airquality)))
colSums(is.na(aq))

vis_miss(aq)

aq_nabular <- nabular(aq)
aq_nabular

aq_nabular |> impute_lm(Ozone ~ Temp)
```

## Collaborative writing
[Trackdown](https://claudiozandonella.github.io/trackdown/)
```{r}
library("trackdown")
```

## Journal templates in quarto
* [Link](https://quarto.org/docs/journals/)
* Can we add a template for Frontiers journals?

## Debugging
* [Link to blog](https://jcarroll.com.au/2023/11/07/print-debugging-now-with-icecream/)
* [Link to package](https://github.com/lewinfox/icecream)
```{r}
library('icecream')
x <- 10
y <- 1:1000
ic(x + y)

ic(cars)

options(icecream.max.lines = 5,
        icecream.prefix = "dbg:",
        icecream.always.include.context = TRUE)
ic(x + y)


options(icecream.peeking.function = utils::head,
        icecream.max.lines = 5,
        icecream.prefix = "dbg:",
        icecream.always.include.context = TRUE)

ic(x + y)
ic(cars)
```

## Update R
* [Link at r-project](https://cloud.r-project.org/bin/macosx/)

## Installed packages
```{r}
dd <- installed.packages()
dd[, 'Package']
dput(dd[, 'Package'], file = paste0('installed_packages_', Sys.Date(), '.RData'))
dput(dd, file = paste0('installed_packages_with_version_', Sys.Date(), '.RData'))
```


## Interesting packages for annoting ggplot plots
[Link](https://jkunst.com/blog/posts/2023-07-10-my-favorite-ggplot2-packages-with-examples/)