---
title: F1NOVF1 with interactions vs. individual t-tests
date: 2023-08-28
categories:
  - R
execute:
  cache: true
---

```{r}
#| label: init
#| message: false
library('tidyverse')
library('multcomp')
library('knitr')
set.seed(1)
```

## Simulation model
* Normally distributed data
* Two-way ANOVA design
  * Factor `F1` with levels `A` and `B`
  * Factor `F2` with levels `A` and `B`
* Means:
  * `F1` = `A`, `F2` = `A` &rarr; $\mu = 0$
  * `F1` = `A`, `F2` = `B` &rarr; $\mu = 1$
  * `F1` = `B`, `F2` = `A` &rarr; $\mu = 1$
  * `F1` = `B`, `F2` = `B` &rarr; $\mu = 3$
* Standard deviation of error:
  * $\sigma = 1$
* In treatment coding, this corresponds to the coefficient vector $\beta = (0, 1, 1, 1)^T$

  
## Example of simulated data
```{r}
generate_data <- function(n = 10, beta = c(0, 1, 1, 1)) {
  dd <- expand_grid(id = 1:n, F1 = c('A', 'B'), F2 = c('A', 'B'))
  X <- model.matrix(~ F1 * F2, data = dd)
  dd |> mutate(y = (X %*% beta)[, 1] + rnorm(nrow(dd)))
}

generate_data() |> 
  ggplot(aes(y = y)) + 
  geom_boxplot() + 
  facet_wrap(vars(F1, F2), labeller = label_both, ncol = 4) +
  theme(axis.text.x=element_blank(),
        axis.ticks.x=element_blank())
```

## Simulate many datasets in order to estimate power to test the hypotheses
* For ANOVA analysis the multcomp package is used in order to simulatneously test the hypotheses (correction for multiple testing with respect to the three hypotheses is included by design)
* Calculate p-values testing the main effects ($\beta_1 = 0$ and $\beta_2 = 0$) and the interaction ($\beta_3 = 0$) for each of the simulations
* Estimate power empirically

### Power of ANOVA analysis
```{r}
simulate_two_way_ANOVA <- function() {
  fm <- lm(y ~ F1 * F2, data = generate_data())
  p_values <- summary(glht(fm, linfct = paste(names(coef(fm))[-1], '= 0')))$test$pvalues
  attributes(p_values) <- NULL
  return(p_values)
}

nr_of_sims <- 1e4
tibble(Effect = c('Main effect 1', 'Main effect 2', 'Interaction'),
       Power = rowSums(replicate(n = nr_of_sims, simulate_two_way_ANOVA()) < 
                         0.05) / nr_of_sims) |> 
  kable(align = 'c', digits = 2)
```

### Power of separate t-tests
```{r}
power.t.test(n = 10, delta = 1, sd = 1, sig.level = 0.05 / 2)
```

### Conclusion
* Power of multivariate test with three hypotheses is approximately equal to power of two separate t-tests with group size $n = 10$
* In ANOVA setting with additional group, the interaction can be tested (power is a little less)
* Note that in the given scenario, 4 groups à 10 individuals are needed in the ANOVA case and 3 groups à 10 individuals are needed in the case of separate t-tests (provided that the control group is 'recycled')

## Appendix
### Use `power.anova.test` to replicate the results of `power.t.test`
```{r}
power.t.test(delta = sqrt(2) * 2 , sd = 2, sig.level = 0.05, power = 0.8)
power.anova.test(groups = 2, between.var = 4, within.var = 4, sig.level = 0.05, power = 0.8)

power.t.test(delta = 2, sd = 1, sig.level = 0.05, power = 0.8)
power.anova.test(groups = 2, between.var = 8, within.var = 4, sig.level = 0.05, power = 0.8)
```