---
title: Kaplan Meier plots
date: 2023-08-03
categories:
  - R
---

Goal: draw facetted Kaplan-Meier-plots with p-value

## With ggquickeda
```{r}
library('tidyverse')
library('survival')
library('ggquickeda')

text_to_add <- 
  tribble(~ sex, ~ text, ~ x, ~ y, 
          'Male', 'p = 0.1', 3000, 0.95,
          'Female', 'p = 0.03', 3000, 0.95)



colon |> 
  mutate(sex = case_when(
    sex == 1 ~ 'Male',
    sex == 0 ~ 'Female')) |> 
  mutate(perfor = factor(perfor)) |> 
  
  ggplot(aes(time = time, status = status, colour = perfor)) + 
  facet_wrap(vars(sex)) +
  geom_km() + geom_kmticks() + 
  geom_label(mapping = aes(label = text, 
                           x = x, 
                           y = y), 
             data = text_to_add, 
             inherit.aes = FALSE) +
  labs(x = 'Time', y = 'Estimated survival')

```


## With survminer
### Do faceting with survminer
```{r}
library('survminer')
fit <- survfit( Surv(time, status) ~ sex + rx, data = colon )
ggsurvplot_facet(fit, colon, facet.by = c('rx', 'perfor'), 
                 pval = TRUE, ggtheme = theme_grey())
```

### Extract plot object so it can be modified in a more usual ggplot2 style
```{r}
fit <- survfit( Surv(time, status) ~ sex + rx + perfor, data = colon )
gg <- ggsurvplot(fit, data = colon, color = 'perfor')
gg$plot + theme_grey() + labs(color = 'perfor', x = 'Time [days]') + facet_wrap(~ sex + rx, labeller = label_both)
```


```{r}
fit <- survfit( Surv(time, status) ~ sex + perfor, data = colon )
gg <- ggsurvplot(fit, data = colon, color = 'perfor', pval = TRUE, facet.by = 'sex')
```

### Show possible problem in survminer
* After facetting, the p-values seem not be correct anymore
```{r}
fit <- survfit(Surv(time, status) ~ sex + rx, data = colon)
gg <- ggsurvplot(fit, pval = TRUE)
gg
gg$plot + facet_wrap(vars(rx))
gg$plot + facet_wrap(vars(rx, sex))
```

### Try to help someone
* [Link](https://github.com/kassambara/survminer/issues/617)
```{r}
library('survival')
library('tidyverse')
library('survminer')

text_to_add <- 
  tribble(~ sex, ~ text, ~ x, ~ y, 
          0, 'Some text', 1000, 0.1,
          1, 'Some other text', 1000, 0.1)

fit <- survfit(Surv(time, status) ~ sex, data = colon)
gg <- ggsurvplot(fit)
gg$plot + facet_wrap(vars(sex)) + theme_gray() + 
  geom_label(aes(x = x, y = y, label = text),
             text_to_add, inherit.aes = FALSE)
```


```{r}
gg <- ggsurvplot(fit, palette = c('black', 'black'))
gg$plot + facet_wrap(vars(sex)) + theme_gray() + 
  geom_label(aes(x = x, y = y, label = text),
             text_to_add, inherit.aes = FALSE) +
  theme(legend.position = "none")

```

## With ggsurvfit
* Link to package: [Link](https://www.danieldsjoberg.com/ggsurvfit/index.html)
* Link to article: [Link](https://www.danieldsjoberg.com/ggsurvfit/articles/gallery.html)
* I found this package later -- I don't know why I haven't across it earlier
* At first glance, it seams actively maintained and seems to produce really nice plots

```{r}
library('ggsurvfit')
glimpse(df_colon)

survfit2(Surv(time, status) ~ surg, data = df_colon) |>
  ggsurvfit(linewidth = 1) +
  add_confidence_interval() +
  add_risktable() +
  add_quantile(y_value = 0.6, color = "gray50", linewidth = 0.75)
```


```{r}
library('tidyverse')
library('survival')
library('tidycmprsk')
library('ggsurvfit')

n <- 100
dat <- tibble(t_death = rexp(n),
              t_relapse = rexp(n),
              t_censoring = rexp(n),
              t_event = pmin(t_death, t_censoring),
              event = ifelse(t_censoring < t_death, 0, 1),
              group = sample(c('Group A', 'Group B'), replace = TRUE, size = 100))

survfit2(Surv(t_event, event) ~ group, data = dat) |> 
  ggsurvfit() + 
  facet_wrap(~ strata) + 
  add_censor_mark()


cuminc(Surv(ttdeath, death_cr) ~ trt + stage, trial) %>%
  ggcuminc(outcome = "death from cancer") + facet_wrap( ~ strata)


survfit2(Surv(time, status) ~ surg, data = df_colon) %>%
  ggsurvfit() +
  add_confidence_interval() +
  facet_wrap(~strata, nrow = 1) +
  theme(legend.position = "none") +
  scale_x_continuous(n.breaks = 6) +
  labs(title = "PFS by Duration between Surgery and Treatment")
```

