---
title: Generation of *in silico* studies
date: 2023-08-21
draft: true
categories:
  - 'Algorithmic study design'
---


## Description of simulation model
* `disease`: Random names for quantity of interest (in reality something like 'Hba1c', 'Survival after therapy with Imatinibe' etc.)
  * 3 letters, duplicates allowed
* `assumed_delta`: Assumed difference in means that has been used for designing the study, lognormal distribution is assumed
* `actual_delta`: *Ground truth* difference in means, assumed 0 in 50% of the cases, in the rest of the cases `assumed_delta` multiplied by random number uniformly distributed between 0.2 and 1.3
* `sd`: Standard deviation of `disease` - some hopefully reasonable random number
* `n_planned`: Designed group size according to `assumed_delta` and `sd` based on `stats::power.t.test()`
* `n_per_year_assumed`: Assumed mean number of patients that can be included in study
* `n_per_year_actual`: Actual number of patients that can be included per year
* `expected_study duration` and `actual_study_duration`: Designed group size divided by expected / actual number of patients per year
* `actual_power`: Power calculated based on `actual_delta` and `sd`
* `null_rejected`: Boolean variable based on Bernoulli trial with $p$ taken from `actual_power`
* `utiliy_null_rejected`: Utility of study - uniformly distributed random number between 0 and 1
* `utiliy_null_not_rejected`: `utiliy_null_rejected` / 2
* `cost_per_patient`: Monetary cost to include a patient in the study - random number on arbitrary scale



## Simulation results
```{r}
#| include: false
#| label: init
library('tidyverse')
library('knitr')
library('skimr')
```

```{r}
#| label: simulation
#| cache: true

# set parameters:
set.seed(87)
nr_of_studies <- 1e4

# utiliy functions:
get_random_names <- function(n, length = 3) {
  replicate(n = n, paste0(sample(letters, size = length), collapse = ''))
}

get_n_planned <- Vectorize(function(delta, sd, power) {
  power.t.test(delta = delta, sd = sd, power = 0.8)$n |> 
    ceiling()
})

get_actual_power <- Vectorize(function(delta, sd, n) {
  power.t.test(delta = delta, sd = sd, n = n)$power
})


# simulate studies:
studies <- tibble(disease = get_random_names(n = nr_of_studies, length = 2),
                  endpoint = get_random_names(n = nr_of_studies, length = 1),
                  assumed_delta = rlnorm(nr_of_studies),
                  actual_delta = ifelse(runif(nr_of_studies) > 0.5, 
                                        0, 
                                        assumed_delta * runif(nr_of_studies, 0.2, 1.3)),
                  sd = assumed_delta * (1 + rlnorm(nr_of_studies)),
                  n_planned = 
                    get_n_planned(delta = assumed_delta, 
                                  sd = sd, 
                                  power = 0.8),
                  n_per_year_assumed = 10 + rlnorm(nr_of_studies),
                  n_per_year_actual = 
                    n_per_year_assumed * runif(n_planned, 0.5, 1.2),
                  expected_study_duration = n_planned / n_per_year_assumed,
                  actual_study_duration = n_planned / n_per_year_actual,
                  actual_power = get_actual_power(delta = actual_delta, 
                                                  sd = sd, 
                                                  n = n_planned),
                  null_rejected = runif(nr_of_studies) < actual_power,
                  utility_null_rejected = runif(nr_of_studies),
                  utility_null_not_rejected = utility_null_rejected / 2,
                  cost_per_patient = rlnorm(nr_of_studies))
```

## Analysis
### 20 randomly chosen studies:
```{r}
sample_n(studies, 20) |> 
           kable(align = 'c', digits = 2)
```

### Descripive statistics
```{r}
skimr::skim(studies)
```

### Additional aspects:

* `expected_study_duration` and `actual_study_duration` have not very realistic values yet &rarr; maybe not all of the considered studies would have been feasible due to insufficient patient numbers `expected_study_duration` *very* long for some studies







