---
title: Modeling autocorrelation
date: 2026-03-17
categories:
  - R
execute:
  eval: false
---



```{r}
#| label: init
#| include: false
library('tidyverse')
```


## Background
* Information from ChatGPT: [Link](https://chatgpt.com/share/69b91479-dd5c-8013-9c58-f29793f92393)

## Wiener process
* English wikipedia: [Link](https://en.wikipedia.org/wiki/Wiener_process)


```{r}
delta_x <- rnorm(1e3)
x <- cumsum(c(0, delta_x))
plot(x)
```



## Ornstein–Uhlenbeck process

## Modling autocorrelation in my case
* I have a mathematical model y(t) = systematic + error
* I assume that the error terms are correlated: $Cov(\epsilon(t_1), \epsilon(t_2) = \lambda^{t_2 - t_1}$ with $0 > \lambda > 1$
* I want to derive a formular for my loss function that takes autocorrelation into account

## Simulating an auto-correlated time series
```{r}




n <- 10000

x <- seq(0, 200, length.out = n)
y_true <- sin(x)

sig_u <- 0.2
sig_v <- 0.2
lambda <- 0.1

u <- rnorm(n, 0, sd = sig_u)
v <- rnorm(n, 0, sd = sig_v)


for(i in 2:n) {
  rho <- lambda^(x[i] - x[i - 1])
  v[i] <- rnorm(1, 
                rho * v[i - 1], 
                sig_v^2 * (1 - rho^2))
  
}

op <- par(mfrow = c(3, 1)); plot(y_true, type = 'l'); points(y_true + u + v, col = 'red', pch = '.', cex = 5); plot(u); plot(v); par(op)

sd(u)
sd(v)

```




## Challenge
I am trying to simulate error terms $\epsilon = Y_{obs} - \mathbb{E}[Y | t, \theta]$, where $t$ is time and $\theta$ the parameter vector of my model. 

I want to simulate a vector $\epsilon$ with $n$ entries that satisfies the following conditions:


* $\epsilon_i \sim \mathcal{N}(0, \sigma^2)$
* $\mathrm{Cov}(\epsilon_i, \epsilon_{i + 1}) = \lambda^{t_{i + 1} - t_{i}}$, $t$ being another vector with $n$ entries that satisfy $t_i > t_{i-1}$ and $0 < \lambda < 1$

## Ornstein–Uhlenbeck process

Long chatGPT dialogue that contains the code to calculate the likelihood if the error terms are generated by an Ornstein-Uhlenbeck process.




```{r}
sigma <- .3
theta <- 0.0001
y <- rnorm(1, 0, sigma^2  / (2 * theta))
n <- 1e5
for (i in 2:n) {
  y[i] <- y[i - 1] - theta * y[i - 1] + rnorm(1, 0, sigma)
}

plot(y, type = 'l')

```

## Calculate likelihood for Ornstein-Uhlenbeck-process generated residuals
* $\theta$ and $\sigma$ unknown
* Approach:
  * Function that calculates loglik in dependence of $\theta$ and $\sigma$
  * Optimize $\theta$ and $\sigma$
  * Maybe calculate confidence interval for $\theta$ and see if parameters can be retrieved from simulated data?





