---
title: 'How bad are bananas'
author: Sebastian Gerdes
date: '2023-02-27'
bibliography: how_bad_are_bananas.bib
# categories: [testing]
execute: 
  echo: false
---
```{r}
#| label: init
#| message: false
library('readODS')
library('tidyverse')
library('knitr')
dat <- read_ods('raw_data.ods')
dat$short_item <- substr(dat$Item, 1, 40)
```




## Discovery
Recently, I have come a across a book that I find really wonderful: "How bad are bananas" by Mike Berners-Lee @berners-lee_how_2020

I would really like to use some of the data for a school project (of course citing the source in an adequate manner). It would be nice to use the data to develop a simple footprint calculator that is available online. Explicite permission from the author to use the data from his book with appropriate inidication of the data source is pending.

## Table
Here, I started to collect some of the data in the book and put them in a table:
```{r}
dat[, c(1:2)] %>% kable()
```

## Plots
```{r}
#| label: co2e_overview
#| fig.height: 10
dat %>% ggplot(aes(CO2e, short_item)) + 
  geom_point(size = 2) +
  scale_x_continuous(trans = 'log10')
```

## References
