---
title: Specifying mixed effects in lme4
date: 2023-06-08
categories:
  - R
---

## Description of trial
* $n$ individuals ($i = 1, 2, .., n$) are measured $m$ times ($j = 1, 2, ..., m$) pre-intervention (denoted $y_{ij}^{pre}$) and post-intervention (denoted $y_{ij}^{post}$)

## Underlying probability model
### Notation
* $\beta_0$ fixed effect intercept
* $\beta_1$ fixed effect of intervention
* $\epsilon \sim \mathcal{N}(\mu = 0, sd = \sigma_0)$, for reasons of better legibility sub- and superscript are omitted here
* $b_{1,i}$ random effect baseline, $b_{1,i} \sim \mathcal{N}(\mu = 0, sd = \sigma_1)$
* $b_{2,i}$ random effect of intervention, $b_{2,i} \sim \mathcal{N}(\mu = 0, sd = \sigma_2)$

### Parameters of the model to be estimated
* $\beta_0$, $\beta_1$, $\sigma_0$, $\sigma_1$ and $\sigma_2$

### Model equations
* Measurement pre intervention:
\begin{align}
y_{ij}^{pre} = \beta_0 + b_i + \epsilon_{ij}^{pre}
\end{align}
* Measurement post-intervention:
\begin{align}
y_{ij}^{post} = \beta_0 + \beta_1 + b_{1,i} + b_{2,i} + \epsilon_{ij}^{post}
\end{align}

### Simulate data
```{r}
#| label: init
#| include: false
library(tidyverse)
library(lme4)
library(knitr)
set.seed(1)
```


```{r}
# size of sample:
n <- 20
m <- 5

# model parameters:
beta_0 <- 0
beta_1 <- 1
sigma_0 <- 2
sigma_1 <- 1
sigma_2 <- 5

# construct data:
patients <- tibble(patient_id = 1:n, 
                   patient_re = rnorm(n, 0, sigma_1),
                   treatment_re = rnorm(n, 0, sigma_2))

dat <- expand_grid(patients,
                   replicate = 1:m,
                   timepoint = c('pre', 'post'))

dat <- dat |> 
  mutate(y = beta_0 + 
           patient_re + 
           ifelse(timepoint == 'post', treatment_re + beta_1, 0) + 
           rnorm(n * m * 2, 0, sigma_0))

# make timepoint a factor to get order in plots right:
dat <- dat |> 
  mutate(timepoint = factor(timepoint, levels = c('pre', 'post')))
         
```

### Tabular characteristics
```{r}
dat |> 
  group_by(timepoint) |> 
  summarize(mean = mean(y)) |> 
  kable()
```



### Visualize data
```{r}
#| label: visualize
if(n < 1e2) {
  dat |>
    ggplot(aes(timepoint, y, col = factor(patient_id))) +
    geom_jitter(width = 0.1, height = 0) + labs(color = 'Patient ID')
}
```

### Fit model
#### Linear model
```{r}
# linear model to get beta_1 and beta_2 
(fm_lm <- lm(y ~ timepoint, data = dat)) |> summary()
```
#### Mixed effects model **with** correlation between `(Intercept)` and `timepointpost`
```{r}
(fm_lmer <- lmer(y ~ timepoint + (timepoint | patient_id), data = dat))
```

#### Mixed effects model **without** correlation between `(Intercept)` and `timepointpost`
* The specifications I tried seem almost right, but not quite:
```{r}
(fm_lmer <- lmer(y ~ timepoint + (timepoint - 1 || patient_id), data = dat))
(fm_lmer <- lmer(y ~ timepoint + (timepoint - 1 | patient_id), data = dat))
(fm_lmer <- lmer(y ~ timepoint + (1 | patient_id) + (timepoint - 1 | patient_id), data = dat))
(fm_lmer <- lmer(y ~ timepoint + (1 | patient_id) + (timepoint + 0 | patient_id), data = dat))
```

