---
title: 'Interfaces to computer algebra systems in R'
author: Sebastian Gerdes
date: '2023-02-25'
# categories: [testing]
---
Herre, I have collected a few resources in the field of computer algebra systems that have an interface to `R`.

## General
* CRAN task view: Section "Multi-Precision Arithmetic and Symbolic Mathematics" on [this](https://cran.r-hub.io/web/views/NumericalMathematics.html) site
* [Blog article](https://blog.ephorie.de/doing-maths-symbolically-r-as-a-computer-algebra-system-cas)

## Maxima
* [Rim package](https://rcst.github.io/rim/)
  * extended search path according to [these](https://techpp.com/2021/09/08/set-path-variable-in-macos-guide/) instructions
  * posted issue on rim package github site because it is still not working [here](https://github.com/rcst/rim/issues/30)
  * edited /etc/launch.d
  * added shell script /usr/bin/maxima

```{r}
system('export PATH=${PATH}:/Applications/Maxima.app/Contents/Resources/opt/bin')
system('echo $PATH')
library(rim)
help(package='rim')
rim::maxima.isInstalled()
rim::maxima.get("1+1;")

```

##  Ryacas
```{r}
library(Ryacas)
as_r(yac_str("Simplify(a*b*a^2/b-a^3)"))
yac_str('Simplify(x^2 + 2*(x-3)^3)')
yac_str("TeXForm(x^2 - 1)")
```
