---
title: 'Humidity'
author: 'Sebastian Gerdes'
date: '2022-09-08'
# categories: [testing]

---

```{r init}
# devtools::install_github('sebastian-gerdes/smisc')
library('smisc')
library('knitr')
library('tibble') 
library('lattice')
```

```{r}
plot(absolute_humidity, -20, 20, log = 'y')
```

```{r}
volume_of_flat <- 120 * 3
relative_humidity_inside <- 50
relative_humidity_outside <- 70
temperature_inside <- 20
temperature_outside <- 0
(water_in_air_inside <- absolute_humidity(relative_humidity = relative_humidity_inside,
                                  temperature = temperature_inside) * volume_of_flat)

(water_in_air_outside <- absolute_humidity(relative_humidity = relative_humidity_outside,
                                  temperature = temperature_outside) * volume_of_flat)
water_out <- water_in_air_inside - water_in_air_outside
```

Conclusion: Airing once with an indoor temperature of `r temperature_inside` degress Celsius with a relative humidity of `r relative_humidity_inside`% and an outdoor temperature of `r temperature_outside` degrees Celsius with a relative outdoor humidity of `r relative_humidity_outside`% transports approx. `r signif(water_out, 2)` liters of water out of the flat, assuming a volume of 360 $m^3$ of the flat.

```{r fig.height=12, fig.width=12}
temperatures <- seq(from = -10, to = 35)
relative_humidities <- seq(from = 0, to = 100, by = 5)
result <- matrix(nrow = length(temperatures), 
                 ncol = length(relative_humidities), 
                 dimnames = list(temperatures, relative_humidities))
for (i in 1:length(temperatures)) {
  for (j in 1:length(relative_humidities)){
    result[i, j] <- absolute_humidity(temperature = temperatures[i], 
                                      relative_humidity = relative_humidities[j])
  }
}
my_breaks <- c(0.5, 0.7, 1, 1.5, 2, 3, 4, 6, 8, 10, 12, 16, 20, 30) / 1000
contour(result, levels = c(0.5, 1, 2, 3, 4, 10, 20) / 1000)
contourplot(result, at = my_breaks,
            scales=list(x=list(rot=90)), 
            xlab = 'Temperature (degrees Celsius)',
            ylab = 'Relative humidity (%)')

```

## Absolute humidity in 1 $m^3$ at a relative humiditiy of 100%

```{r}
# scenario 1:
temperatures <- seq(from = -10, to = 35, by = 5)
relative_humidities <- seq(from = 0, to = 100, by = 10)
result <- matrix(nrow = length(temperatures), 
                 ncol = length(relative_humidities), 
                 dimnames = list(temperatures, relative_humidities))
for (i in 1:length(temperatures)) {
  for (j in 1:length(relative_humidities)){
    result[i, j] <- absolute_humidity(temperature = temperatures[i], 
                                      relative_humidity = relative_humidities[j])
  }
}

kable(round(result, digits = 2)) 
# scenario 2:
temperatures <- -10:30
abs_humidity <- round(absolute_humidity(temperature = temperatures), digits = 3)
kable(tibble(temperatures = temperatures, abs_humidity = abs_humidity))
```
